R submission Pilots to FDA

### 2022-10-24 slides/r4csr-psi.Rmd
add psi slides

---
title: "R for Clinical Study Reports and Submission"
subtitle: "RStudio Pharma Meetup Series"
author: "Yilong Zhang"
date: "May 2022"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
bibliography: reference.bib
---

@@

# R Consortium pilot submission

## Working group

- The R submission working group is a cross-industry pharma working group focusing on improving practices of R-based clinical trial regulatory submission.
  - [members and affiliations](https://rconsortium.github.io/submissions-wg/)
- Mission: Easier R-based clinical trial regulatory submissions
- [pilot 1 - common analyses](https://rconsortium.github.io/submissions-wg/pilot-overall.html#pilot-1---common-analyses)
  - 2021.11   Initial submission to FDA. 
  - 2022.12   Received FDA response 
  - 2022.02   Revised submission to FDA
  - 2022.03   Received final FDA response.

## Challenges and assumptions

@@

## FDA response

@@

## Future work

- Pilot R Shiny app submission to FDA.
- Seek pilot submission opportunities to other regulator agencies. 
- Pilot submission with advanced analysis methods (e.g., study design, missing data, Bayesian, etc).

@@

## Reproducibility spectrum

@@

## Tools for reporting and submission

Tools:

- [`r2rtf`](https://merck.github.io/r2rtf/): create production-ready tables and figures in RTF format.
- [`pkglite`](https://merck.github.io/pkglite/): represent and exchange R package source code as text files.
- `cleanslate` (under internal validation): create portable R environments.

Bookdown: <https://r4csr.org/>

## r2rtf: create TLFs in RTF format

```{r, out.width="100%", echo=FALSE, fig.align="center"}
knitr::include_graphics("function-summary.png")
```

## pkglite: compact package representations

@@

## cleanslate: portable R environments

- Create a project folder with specific context (`.Rproj`, `.Rprofile`, `.Renviron`)
- Install a specific version of R into the project folder
- Install a specific version of Rtools into the project folder 
(without administrator privileges)

```{r, eval=FALSE}
library("cleanslate")

"portable-project/" %>%
  use_project(repo = "https://url/snapshot/2021-11-20/") %>%
  use_rprofile() %>%
  use_renviron() %>%
  use_r_version(version = "4.1.1") %>%
  use_rtools(version = "rtools40")
```

## Summary of workflow as user stories

Within a regulatory R environment:

- As statisticians, we use `tidyverse`, `r2rtf`,
and internal tools to define the mock-up table, listing and figure (TLFs) for statistical analysis of a clinical trial.
  - More than 100 TLFs for efficacy and safety of a drug or vaccine.

- As programmers, we use `tidyverse`, `r2rtf`, and internal tools to develop and/or validate analysis results based on mock-up TLFs.

- As statisticians/programmers, we use `pkglite` and internal tools to prepare proprietary R packages and analysis R scripts for eCTD submission packages.

- As an internal/external reviewers, we use `cleanslate` to re-construct a portable environment (if required) to reproduce analysis results.

More details: <https://r4csr.org/>

## Folder structure

@@

## Cross-industry collaborations

@@

## Future directions

>>>>

---
title: "R submission Pilots to FDA"
subtitle: "PSI Webinar Series: Showcasing R use in Pharma"
author: "Yilong Zhang on Behalf of R Consortium “R Submission” Working Group"
date: "Oct. 2022"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
bibliography: reference.bib
---

@@

# R Consortium Pilot Submission

## The R Consortium R Submission Working Group

- The R submission working group is a cross-industry pharma working group focusing on improving practices of R-based clinical trial regulatory submission.
  - [members and affiliations](https://rconsortium.github.io/submissions-wg/)

```{r, out.width = "60%", echo = FALSE, fig.align = "center"}
knitr::include_graphics("submission-group-logo.png")
```

## Objective and Timeline

- Objective: Create open source example for R-based clinical trial regulatory submissions
- [pilot 1 - common analyses](https://rconsortium.github.io/submissions-wg/pilot-overall.html#pilot-1---common-analyses)
  - 2021.11   Initial submission to FDA. 
  - 2022.12   Received FDA response 
  - 2022.02   Revised submission to FDA
  - 2022.03   Received final FDA response.

- [pilot 2 - shiny app](https://github.com/RConsortium/submissions-pilot2-to-fda)
  - 2022.10   Public review of initial submission to FDA.

- [pilot 3 - ADaM data](https://github.com/RConsortium/submissions-wg/blob/main/Documents/R%20Submission%20%20Pilot%201%20Extension%20Proposal.pdf)
  - 2022.10   Proposal initiation.

## Pilot 1 Challenges and Assumptions

@@

## FDA Response

@@

## Future Work

- Seek pilot submission opportunities to other regulator agencies. 
- Pilot submission with advanced analysis methods (e.g., study design, missing data, Bayesian, etc).

@@

## Reproducibility Spectrum

@@

## Tools for Submission

- Tool: [`pkglite`](https://merck.github.io/pkglite/), represent and exchange R package source code as text files.

- Reference: @zhao2022electronic

- Book: <https://r4csr.org/>

## pkglite: Compact Package Representations

@@

## Folder Structure

@@

## Cross-industry Collaborations

@@

## Future Directions
